Synthesising Recursive Functions for First-Order Model Counting

### 2023-02-05 doc/paper/paper.tex
2/4 proofs complete (an overhaul of Preliminaries)

\title{Recursive Solutions to First-Order Model Counting}

% TODO: double check how the university names should be formatted and the emails
% TODO: read the formatting instructions and their LaTeX source
% TODO: make sure there are no overflow errors
% TODO: for arbitrary instances of '2022' (e.g., in citations), make it '2023'
% TODO: maybe mention the term 'data complexity'
% TODO: fix doom emacs warnings
\author{%
Paulius Dilkas$^1$\and
Vaishak Belle$^2$\\
\affiliations
$^1$National University of Singapore, Singapore\\
$^2$University of Edinburgh, UK\\
\emails
paulius.dilkas@nus.edu.sg,
vbelle@ed.ac.uk
}

\crefname{algocf}{algorithm}{Algorithms}
\Crefname{algocf}{Algorithm}{Algorithms}

\newcommand{\crefrangeconjunction}{--}
\newcommand\pfun{\mathrel{\ooalign{\hfil$\mapstochar\mkern5mu$\hfil\cr$\to$\cr}}}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\newcommand{\FOtwo}{$\mathsf{FO}^{2}$}
\newcommand{\FOthree}{$\mathsf{FO}^{3}$}
\newcommand{\SFO}{$\mathsf{S}^{2}\mathsf{FO}^{2}$}
\newcommand{\SRU}{$\mathsf{S}^{2}\mathsf{RU}$}
\newcommand{\Uone}{$\mathsf{U}_{1}$}
\newcommand{\Ctwo}{$\mathsf{C}^{2}$}
\newcommand{\IFO}{$\mathsf{I}\mathsf{FO}^{2}$}
\DeclareMathOperator{\CR}{\textsc{CR}}
\DeclareMathOperator{\GDR}{\textsc{GDR}}
\DeclareMathOperator{\Reff}{\textsc{Ref}}
\DeclareMathOperator{\dom}{dom}
\DeclareMathOperator{\Doms}{Doms}
\DeclareMathOperator{\MC}{MC}
\DeclareMathOperator{\Vars}{Vars}
\SetKwProg{Fn}{Function}{:}{}
\SetKwFunction{identifyRecursion}{r}
\SetKwFunction{generateMaps}{genMaps}
\makeatletter
\newcommand{\nosemic}{\renewcommand{\@endalgocfline}{\relax}}% Drop semi-colon ;
\newcommand{\dosemic}{\renewcommand{\@endalgocfline}{\algocf@endline}}% Reinstate semi-colon ;
\newcommand{\pushline}{\Indp}% Indent
\newcommand{\popline}{\Indm\dosemic}% Undent
\makeatother
\Crefname{algocf}{Algorithm}{Algorithms}
\crefname{line}{line}{lines}

@@

@@

In this work, we show how to relax these restrictions in a way that results in a
stronger (W)FOMC algorithm, capable of handling more instances (e.g., counting
various injective mappings) in a lifted manner. The ideas presented in this
paper are implemented in \textsc{Crane}---an extension of the arguably most
well-known WFOMC algorithm \textsc{ForcLift}. \textsc{ForcLift} works in two
stages: compilation and evaluation/propagation. In the first stage, various
\emph{(compilation) rules} are applied to the input (or some derivative)
formula, gradually constructing a circuit. In the second stage, the weights of
the instance are propagated through the circuit, computing the weighted model
count. Along with new compilation rules, \textsc{Crane} introduces changes to
both stages of the process. First, while \textsc{ForcLift} applies compilation
rules via greedy\footnote{The algorithm is not described in any paper but can be
  found in its source code at \url{https://dtai.cs.kuleuven.be/drupal/wfomc}}
search, \textsc{Crane} also supports a hybrid search algorithm that applies some
rules greedily and some using breadth-first search. Second, the product of
compilation is not directly evaluated but rather interpreted as a collection of
functions on domain sizes.

@@

The main conceptual difference between \textsc{Crane} and \textsc{ForcLift} is
that we utilise labelled directed graphs instead of circuits, where cycles
represent recursive calls. Suppose the input formula $\phi$ depends on a domain
of size $n \in \mathbb{N}$. \emph{Generalised domain recursion} (GDR)---one of
the new compilation rules---transforms $\phi$ into a different formula $\psi$
that has an additional constant and some new \emph{constraints}. After some
additional transformations, the constraints in $\psi$ become `uniform' and can
be removed, replacing the domain of size $n$ with a new domain of size
$n-1$---this is the responsibility of the \emph{constraint removal} (CR)
compilation rule. Afterwards, another compilation rule recognizes that the
resulting formula is just like the input formula $\phi$ but with a different
domain. This observation allows us to add a cycle-forming edge to the graph,
which can be interpreted as function $f$ relying on $f(n-1)$ to compute $f(n)$.

@@

Our representation of FOMC instances is largely based on the format used
internally by \textsc{ForcLift}, some aspects of which are described by Van den
Broeck et al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11}. \textsc{ForcLift} can
translate sentences in a variant of function-free many-sorted FOL with equality
to this internal format. We use lowercase Latin letters for predicates and
constants, uppercase Latin letters for variables, and uppercase Greek letters
for domains. An \emph{atom} is $p(t_1, \dots, t_n)$ for some predicate $p$ and
terms $t_{1}, \dots, t_{n}$. A \emph{term} is either a constant or a variable. A
\emph{literal} is either an atom or the negation of an atom (denoted by
$\neg p(t_1, \dots, t_n)$). Let $\mathcal{D}$ be the set of all (finite)
domains. Initially, $\mathcal{D}$ contains all domains mentioned by the input
(W)FOMC instance. During compilation, new domains are added to $\mathcal{D}$ by
some of the compilation rules. Each such new domain is interpreted as a subset
of some element of $\mathcal{D}$.

@@

Similarly to variables in \cref{def:clause}, all constants are (implicitly)
mapped to domains, and each $n$-ary predicate is implicitly mapped to a sequence
of $n$ domains. For constant or variable $x$, predicate $p$, and domains
$\Gamma$ and $\Delta$, we write, e.g., $x \in \Gamma$ and
$p \in \Gamma \times \Delta$ to denote that $x$ is associated with $\Gamma$, and
$p$ is associated with $\Gamma$ and $\Delta$ (in that order).

\begin{definition}[Formula]\label{def:formula}
  A \emph{formula} (called a c-theory by Van den Broeck et
  al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11}) is a set of clauses s.t.\ all
  constraints and atoms `type check' with respect to domains.
\end{definition}

\begin{example}\label{example:first}
  Let $\phi \coloneqq \{\, c_1, c_2 \,\}$ be a formula with clauses
  \begin{align*}
    c_1 &\coloneqq
          \begin{multlined}[t]
            (\{\, \neg p(X, Y), \neg p(X, Z) \,\}, \{\, (Y, Z) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Delta \,\}),
          \end{multlined}\\
    c_2 &\coloneqq
          \begin{multlined}[t]
            (\{\, \neg p(X, Y), \neg p(Z, Y) \,\}, \{\, (X, Z) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Gamma \,\})
          \end{multlined}
  \end{align*}
  for some predicate $p$, variables $X$, $Y$, $Z$, and domains $\Gamma$ and
  $\Delta$. Then $\Vars(\{\, (Y, Z) \,\}) = \{\, Y, Z \,\}$, and
  $\Vars(c_{1}) = \Vars(c_{2}) = \{\, X, Y, Z \,\}$. Based on the domain maps of
  $c_{1}$ and $c_{2}$, we can infer that $p \in \Gamma \times \Delta$. All
  variables (in both clauses) that occur as the first argument to $p$ are in
  $\Gamma$, and, likewise, all variables that occur as the second argument to
  $p$ are in $\Delta$. Therefore, $\phi$ `type checks' as a valid formula.
\end{example}

There are two major differences between
\cref{def:constraint,def:clause,def:formula} and the corresponding concepts
introduced by Van den Broeck et al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11}.
First, we decouple variable-to-domain assignments from constraints and move them
to a separate function $\delta_{c}$ in \cref{def:clause}. Formalising these
assignments as a function unveils the (previously implicit) assumption that each
variable must be assigned to a domain. Second, while Van den Broeck et
al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11} allow for equality constraints and
constraints of the form $X \not\in \Delta$ for some variable $X$ and domain
$\Delta$, we exclude such constraints simply because they are not needed.

One can read a formula from \cref{def:formula} as a sentence in a FOL\@. All
variables in a clause are implicitly universally quantified (but note that
variables are never shared among clauses), and all clauses in a formula are
implicitly linked by a conjunction. Thus, formula $\phi$ from
\cref{example:first} reads as
\begin{align*}
  &\begin{multlined}[t]
    (\forall X \in \Gamma\text{. }\forall Y, Z \in \Delta\text{. }\\
    Y \ne Z \implies \neg p(X, Y) \lor \neg p(X, Z)) \land
    \end{multlined}\\
  &\begin{multlined}[t]
    (\forall X, Z \in \Gamma\text{. }\forall Y \in \Delta\text{. }\\
    X \ne Z \implies \neg p(X, Y) \lor \neg p(Z, Y)).
    \end{multlined}
\end{align*}

Once domains are mapped to finite sets and constants to specific (and different)
elements in those sets, a formula can be viewed as a set of conditions that the
predicates (interpreted as relations) have to satisfy. Hence, FOMC is the
problem of counting the number of combinations of relations that satisfy these
conditions.

\begin{example}
  Let $\phi$ be as in \cref{example:first} and let $|\Gamma| = |\Delta| = 2$.
  There are $2^{2 \times 2} = 16$ possible relations between $\Gamma$ and
  $\Delta$. Let us count how many of them satisfy the conditions imposed on
  predicate $p$. The empty relation does. All four relations of cardinality one
  do too. Finally, there are two relations of cardinality two that satisfy the
  conditions as well. Thus, the FOMC of $\phi$ (when $|\Gamma| = |\Delta| = 2$)
  is 7. Incidentally, the FOMC of $\phi$ counts partial injections. We will
  continue to use the problem of counting partial injections (and the formula
  from \cref{example:first} specifically) as the main running example throughout
  the paper.
\end{example}
% TODO: It might be useful to give one or two readable examples with interesting
% model counts. For example Px v Qx over domain n has 3^n solutions. I think
% The Frends smokers example has a more complicated solution you can find in one
% of the papers. It might motivate the framework generally speaking better
% although of course you can continue to use the same running example for the
% rest of your paper. Such examples are especially useful for readers who are
% not familiar but still need to get excited by the problem

\paragraph*{Notation.}
We write $\langle\rangle$ and $\langle x \rangle$ to denote an empty list and a
list with one element $x$, respectively, and $|l|$ for the length of list $l$.
We write $\pfun$ for partial functions. Let $S$ be a set of constraints or
literals, $V$ a set of variables, and $x$ a variable or a constant. We write
$S[x/V]$ to denote $S$ with all occurrences of all variables in $V$ replaced
with $x$.

@@

\paragraph{Notation.}
Let $\Doms$ be a function that maps any clause or formula to the set of domains
used within, and let $\dom(f)$ denote the domain of function $f$.

@@

  \ForEach{formula and node $(\psi, v) \in C(\#\phi)$}{
    $\rho \gets \identifyRecursion{$\phi$, $\psi$}$\;
    \lIf{$\rho \ne {\normalfont \texttt{null}}$}{\Return{$\{\, (\Reff_\rho(v), \langle\rangle) \,\}$}}
  }
  \Return{$\emptyset$}\;

@@

% TODO: maybe something other than MC? Introduce the notation and the role of \sigma.
% TODO: add an argument about why this recursion is (or can easily be made) finite
\begin{theorem}[Correctness of $\Reff$]
\end{theorem}
\begin{proof} % TODO: finish (Proof technique: show a bijection between the sets of models)
  For a given formula $\phi$, \cref{alg:trycache} attempts to find a formula
  $\psi$ and a map $\rho\colon \Doms(\psi) \to \Doms(\phi)$ such that
  $\MC(\psi, \sigma \circ \rho\colon \Doms(\psi) \to \mathbb{N}_{0}) = \MC(\phi, \sigma\colon \Doms(\phi) \to \mathbb{N}_{0})$.
\end{proof}

@@

\begin{theorem}
  Any collection of functions returned by \textsc{Crane} can be evaluated in
  time polynomial in domain sizes.
\end{theorem}
\begin{proof}% TODO (I have some handwritten notes)
  % prove that all functions output by Crane can be computed in polytime (must, should be easy: 1) bound the number of possible arguments that any function can have, 2) prove that the computation time of any dynamic programming cell is always polytime (what about functions depending on other functions?))
\end{proof}

\section{Empirical Results}\label{sec:results}
% TODO: Add that the linear order axiom is also liftable. Cite the same paper in the
% introduction as well.

@@

% TODO: the correctness of all of these was confirmed by finding the right base
% cases and checking whether the function definitions produce the right values
% for a sufficiently high number of domain sizes.

% TODO: in the acknowledgements, mention that most of the work was done at UoE
% TODO: should the appendix be in the same file or not?
% TODO: also mention that most of the work for this paper was done at UoE

>>>>

\title{Synthesising Recursive Functions for First-Order Model Counting}

% TODO: read the formatting instructions and their LaTeX source (email font?)
% TODO: make sure there are no overflow errors
% TODO: maybe mention the term 'data complexity'
% TODO: make claim numbers subordinate to theorem numbers
% TODO: mention that most of the work for this paper was done at UoE
% TODO: the appendix should be in a separate file
% TODO: cite some works about synthesis
% TODO: clean up unnecessary things from the header (e.g., packages and definitions)
% TODO: don't use the '\in' syntax for predicates
% TODO: make sure I'm using \sigma instead of || everywhere
% TODO: (at the end): grammarly the whole thing, put the appendix in a separate file
\author{%
Paulius Dilkas$^1$\and
Vaishak Belle$^2$\\
\affiliations
$^1$National University of Singapore, Singapore, Singapore\\
$^2$University of Edinburgh, Edinburgh, UK\\
\emails
paulius.dilkas@nus.edu.sg,
vbelle@ed.ac.uk
}

\crefalias{diagram}{equation}
\crefname{diagram}{Diagram}{Diagrams}
\Crefname{diagram}{Diagram}{Diagrams}
\creflabelformat{diagram}{#2\textup{(#1)}#3}

\crefname{algocf}{Algorithm}{Algorithms}
\Crefname{algocf}{Algorithm}{Algorithms}
\crefname{condition}{Condition}{Conditions}
\Crefname{condition}{Condition}{Conditions}

\newcommand{\crefrangeconjunction}{--}
\newcommand\pfun{\mathrel{\ooalign{\hfil$\mapstochar\mkern5mu$\hfil\cr$\to$\cr}}}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\newcommand{\FOtwo}{$\mathsf{FO}^{2}$}
\newcommand{\FOthree}{$\mathsf{FO}^{3}$}
\newcommand{\SFO}{$\mathsf{S}^{2}\mathsf{FO}^{2}$}
\newcommand{\SRU}{$\mathsf{S}^{2}\mathsf{RU}$}
\newcommand{\Uone}{$\mathsf{U}_{1}$}
\newcommand{\Ctwo}{$\mathsf{C}^{2}$}
\newcommand{\IFO}{$\mathsf{I}\mathsf{FO}^{2}$}
\DeclareMathOperator{\CR}{\textsc{CR}}
\DeclareMathOperator{\GDR}{\textsc{GDR}}
\DeclareMathOperator{\Reff}{\textsc{Ref}}
\DeclareMathOperator{\dom}{dom}
\DeclareMathOperator{\Doms}{Doms}
\DeclareMathOperator{\Vars}{Vars}
\DeclareMathOperator{\Preds}{Preds}
\SetKwProg{Fn}{Function}{:}{}
\SetKwFunction{identifyRecursion}{r}
\SetKwFunction{generateMaps}{genMaps}
\makeatletter
\newcommand{\nosemic}{\renewcommand{\@endalgocfline}{\relax}}% Drop semi-colon ;
\newcommand{\dosemic}{\renewcommand{\@endalgocfline}{\algocf@endline}}% Reinstate semi-colon ;
\newcommand{\pushline}{\Indp}% Indent
\newcommand{\popline}{\Indm\dosemic}% Undent
\makeatother
\Crefname{algocf}{Algorithm}{Algorithms}
\crefname{line}{line}{lines}

@@

% TODO: In the introduction it is written: "Crane uses a hybrid search algorithm
% that applies some rules greedily and some using breadth-first search". Which
% rules are applied greedily and which are not? What is the reason for parting
% from the usual greedy method?

TODO: cite \textsc{Magician}~\cite{DBLP:conf/aaai/VenugopalSG15} and
\textsc{Tuffy}~\cite{DBLP:journals/pvldb/NiuRDS11} (read the abstracts first,
look at the code)

@@

% TODO: I am not sure the statement of note 2 is correct since, in [van den
% Broeck, IJCAI 2011], the compilation is described as follows: "Compiling a set
% of c-clauses C to Fo d-DNNF is performed by the Compile(C) function, which
% tries to recursively apply the compilation rules in the order we present
% here". But then maybe the algorithm was not yet ForcLift.

In this work, we show how to relax these restrictions in a way that results in a
stronger (W)FOMC algorithm, capable of handling more instances (e.g., counting
various injective mappings) in a lifted manner. The ideas presented in this
paper are implemented in \textsc{Crane}---an extension of the arguably most
well-known WFOMC algorithm \textsc{ForcLift}. \textsc{ForcLift} works in two
stages: compilation and evaluation/propagation. In the first stage, various
\emph{(compilation) rules} are applied to the input (or some derivative)
formula, gradually constructing a circuit. In the second stage, the weights of
the instance are propagated through the circuit, computing the weighted model
count. Along with new compilation rules, \textsc{Crane} introduces changes to
both stages of the process. First, while \textsc{ForcLift} applies compilation
rules via greedy\footnote{The algorithm is not described in any paper but can be
  found in its source code at \url{https://dtaid.cs.kuleuven.be/wfomc}}
search, \textsc{Crane} also supports a hybrid search algorithm that applies some
rules greedily and some using breadth-first search. Second, the product of
compilation is not directly evaluated but rather interpreted as a collection of
functions on domain sizes.

@@

The main conceptual difference between \textsc{Crane} and \textsc{ForcLift} is
that we utilise labelled directed graphs instead of circuits, where cycles
represent recursive calls. Suppose the input formula $\phi$ depends on a domain
of size $n \in \mathbb{N}_{0}$. \emph{Generalised domain recursion} (GDR)---one
of the new compilation rules---transforms $\phi$ into a different formula $\psi$
that has an additional constant and some new \emph{constraints}. After some
additional transformations, the constraints in $\psi$ become `uniform' and can
be removed, replacing the domain of size $n$ with a new domain of size
$n-1$---this is the responsibility of the \emph{constraint removal} (CR)
compilation rule. Afterwards, another compilation rule recognizes that the
resulting formula is just like the input formula $\phi$ but with a different
domain. This observation allows us to add a cycle-forming edge to the graph,
which can be interpreted as function $f$ relying on $f(n-1)$ to compute $f(n)$.

@@

Our representation of FOMC instances is largely based on the format used
internally by \textsc{ForcLift}, some aspects of which are described by Van den
Broeck et al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11}. \textsc{ForcLift} can
translate sentences in a variant of function-free many-sorted FOL with equality
to this internal format. We use lowercase Latin letters for predicates and
constants, uppercase Latin letters for variables, and uppercase Greek letters
for domains. Sometimes we write predicate $p$ as $p/n$, where
$n \in \mathbb{N}^{+}$ is the \emph{arity} of $p$. An \emph{atom} is
$p(t_1, \dots, t_n)$ for some predicate $p/n$ and terms $t_{1}, \dots, t_{n}$. A
\emph{term} is either a constant or a variable. A \emph{literal} is either an
atom or the negation of an atom (denoted by $\neg p(t_1, \dots, t_n)$). Let
$\mathcal{D}$ be the set of all relevant (finite) domains. Initially,
$\mathcal{D}$ contains all domains mentioned by the input formula. During
compilation, new domains are added to $\mathcal{D}$ by some of the compilation
rules. Each such new domain is interpreted as a subset of some other domain in
$\mathcal{D}$.

@@

\paragraph{Notation.}
For any set $S$, let $2^{S}$ denote its power set and
$S^{\ast} \coloneqq \bigcup_{i=0}^{\infty} S^{i}$ the set of tuples of any
finite length of elements of $S$. For any $n \in \mathbb{N}_{0}$, let
$[n] \coloneqq \{\, 1, \dots, n \,\}$. For example, $[0] = \emptyset$, and
$[2] = \{\, 1, 2 \,\}$. We write $\langle\rangle$ for an empty list,
$\langle x \rangle$ for a list with one element $x$, $|l|$ for the length of
list $l$, and $\pfun$ to denote partial functions. Let $S$ be a set of
constraints or literals, $V$ a set of variables, and $x$ a variable or a
constant. We write $S[x/V]$ to denote $S$ with all occurrences of all variables
in $V$ replaced with $x$. Let $\Doms$ (respectively, $\Preds$) be the function
that maps any clause or formula to the set of domains (respectively, predicates)
used within, and let $\dom(f)$ denote the domain of function $f$.

\begin{definition}[Formula]\label{def:formula}
  A \emph{formula} (called a c-theory by Van den Broeck et
  al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11}) is a set of clauses s.t.\ all
  constraints and atoms `type check' with respect to domains. Each formula
  $\phi$ has a map $\pi_{\phi}\colon \Preds(\phi) \to {\Doms(\phi)}^{\ast}$ such
  that, for each predicate $p/n \in \Preds(\phi)$, we have that
  $\pi_{\phi}(p) \in {\Doms(\phi)}^{n}$.\footnote{Similarly to variables and
    predicates, constants are also mapped to domains, although we keep this
    mapping implicit.}
\end{definition}

\begin{example}\label{example:first}
  Let $\phi \coloneqq \{\, c_1, c_2 \,\}$ be a formula with clauses
  \begin{align*}
    c_1 &\coloneqq
          \begin{multlined}[t]
            (\{\, \neg p(X, Y), \neg p(X, Z) \,\}, \{\, (Y, Z) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Delta \,\}),
          \end{multlined}\\
    c_2 &\coloneqq
          \begin{multlined}[t]
            (\{\, \neg p(X, Y), \neg p(Z, Y) \,\}, \{\, (X, Z) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Gamma \,\})
          \end{multlined}
  \end{align*}
  for some predicate $p/2$, variables $X$, $Y$, $Z$, and domains $\Gamma$ and
  $\Delta$. Based on $\delta_{c_{1}}$ and $\delta_{c_{2}}$, we can infer that
  $\pi_{\phi} = \{\, p \mapsto (\Gamma, \Delta) \,\}$. All variables that occur
  as the first argument to $p$ are in $\Gamma$, and, likewise, all variables
  that occur as the second argument to $p$ are in $\Delta$. Therefore, $\phi$
  `type checks' as a valid formula.
\end{example}

There are two major differences between
\cref{def:constraint,def:clause,def:formula} and the corresponding concepts
introduced by Van den Broeck et al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11}.
First, we decouple variable-to-domain assignments from constraints and move them
to a separate function $\delta_{c}$ in \cref{def:clause}. Second, while Van den
Broeck et al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11} allow for equality
constraints and constraints of the form $X \not\in \Delta$ for some variable $X$
and domain $\Delta$, we exclude such constraints simply because they are not
needed.

One can read a formula in our format as a sentence in FOL\@. All variables in a
clause are implicitly universally quantified (but note that variables are never
shared among clauses), and all clauses in a formula are implicitly linked by a
conjunction. Thus, formula $\phi$ from \cref{example:first} reads as
\begin{align*}
  &\begin{multlined}[t]
    (\forall X \in \Gamma\text{. }\forall Y, Z \in \Delta\text{. }\\
    Y \ne Z \implies \neg p(X, Y) \lor \neg p(X, Z)) \land
    \end{multlined}\\
  &\begin{multlined}[t]
    (\forall X, Z \in \Gamma\text{. }\forall Y \in \Delta\text{. }\\
    X \ne Z \implies \neg p(X, Y) \lor \neg p(Z, Y)).
    \end{multlined}
\end{align*}

Once domains are mapped to finite sets and constants to elements in those sets,
a formula can be viewed as a set of conditions that the predicates (interpreted
as relations) have to satisfy. Hence, FOMC is the problem of counting the number
of combinations of relations that satisfy these conditions. To formalise this
idea, let $\sigma\colon \mathcal{D} \to \mathbb{N}_{0}$ be the \emph{domain size
  function} that maps each domain to a non-negative integer.

\begin{definition}\label{def:model}
  Let $\phi$ be a formula and $\sigma$ a domain size function. A \emph{model} of
  $(\phi, \sigma)$ is a map
  $\mathfrak{M}\colon \Preds(\phi) \to 2^{\mathbb{N}_{0}^{\ast}}$ such that the
  following two conditions are satisfied.
  \begin{enumerate}
    \item Let $p/n \in \Preds(\phi)$ be a predicate and let
          $\pi_{\phi}(p) = {(d_{i})}_{i=1}^{n}$ for some domains
          $d_{i} \in \Doms(\phi)$. Then
          \begin{equation}\label{eq:subset}
            \mathfrak{M}(p) \subseteq \prod_{i=1}^{n}[\sigma(d_{i})].\footnote{For simplicity, \cref{def:model} ignores constants. To include constants, one would replace $[\sigma(d_{i})]$ with a set that contains all constants associated with domain $d_{i}$, extended with enough new elements to make its cardinality $\sigma(d_{i})$.}
          \end{equation}
    \item As a collection of relations, $\mathfrak{M}$ satisfies
          $\phi$.
  \end{enumerate}
\end{definition}

\begin{example}
  Let $\phi$ be as in \cref{example:first} and let
  $\sigma(\Gamma) = \sigma(\Delta) = 2$. Then the right-hand side of
  \cref{eq:subset} becomes
  \[
    [\sigma(\Gamma)] \times [\sigma(\Delta)] = [2] \times [2] = \{\,(1, 1), (1, 2), (2, 1), (2, 2)\,\}
  \]
  There are $2^{4} = 16$ possible relations between $\Gamma$ and $\Delta$. Let
  us count how many of them satisfy the conditions imposed on predicate $p$. The
  empty relation does. All four relations of cardinality one do too. Finally,
  there are two relations of cardinality two that satisfy the conditions as
  well. Thus, the FOMC of $(\phi, \sigma)$ is 7. Incidentally, it counts partial
  injections. We will continue to use the problem of counting partial injections
  (and the formula from \cref{example:first} specifically) as the main running
  example throughout the paper.
\end{example}
% TODO: It might be useful to give one or two readable examples with interesting
% model counts. For example Px v Qx over domain n has 3^n solutions. I think
% The Frends smokers example has a more complicated solution you can find in one
% of the papers. It might motivate the framework generally speaking better
% although of course you can continue to use the same running example for the
% rest of your paper. Such examples are especially useful for readers who are
% not familiar but still need to get excited by the problem

@@

@@

  \ForEach{formula and node $(\psi, v) \in C(\#\phi)$}{\label{line:selectformula}
    $\rho \gets \identifyRecursion{$\phi$, $\psi$}$\;
    \lIf{$\rho \ne {\normalfont \texttt{null}}$}{\Return{$\{\, (\Reff_\rho(v), \langle\rangle) \,\}$}}\label{line:rho}
  }
  \Return{$\emptyset$}\;

@@

\begin{theorem}[Correctness of $\Reff$]
  Let $\phi$ be the formula used as input to \cref{alg:trycache}. Let $\psi$ be
  any formula selected on \cref{line:selectformula} of the algorithm such that
  $\rho \ne {\normalfont\texttt{null}}$ on \cref{line:rho}. Then the set of
  models of $(\psi, \sigma \circ \rho)$ is equal to the set of models of
  $(\phi, \sigma)$.
\end{theorem}
\begin{proof}
  We first show that the right-hand side of \cref{eq:subset} is the same for
  models of both $(\phi, \sigma)$ and $(\psi, \sigma \circ \rho)$.
  \Cref{alg:trycache} ensures that $\psi \equiv \phi$ up to domains. In
  particular, this means that $\Preds(\psi) = \Preds(\phi)$. The square in
  \begin{equation}\label[diagram]{diagram:ref}
    \begin{tikzcd}
      \Preds(\psi) \ar[equal]{r} \ar[d, swap, "\pi_{\psi}"] & \Preds(\phi) \ar[d, "\pi_{\phi}"] \\
      {\Doms(\psi)}^{\ast} \ar[r, "\rho"] & {\Doms(\phi)}^{\ast} \ar[r, "\sigma"] & \mathbb{N}_{0}^{\ast}
    \end{tikzcd}
  \end{equation}
  commutes by \cref{def:formula} and the definition of $\rho$. In other words,
  for any predicate $p/n \in \Preds(\psi) = \Preds(\phi)$, function $\rho$
  translates the domains associated with $p$ in $\psi$ to the domains associated
  with $p$ in $\phi$. Let $\pi_{\phi}(p) = {(d_{i})}_{i=1}^{n}$ and
  $\pi_{\psi}(p) = {(e_{i})}_{i=1}^{n}$ for some domains $d_{i} \in \Doms(\phi)$
  and $e_{i} \in \Doms(\psi)$. Since $\rho(e_{i}) = d_{i}$ for all
  $i = 1, \dots, n$ by the commutativity in \cref{diagram:ref}, we get that
  \begin{equation}\label{eq:equality}
    \prod_{i=1}^{n}[\sigma(d_{i})] = \prod_{i=1}^{n}[\sigma \circ \rho(e_{i})]
  \end{equation}
  as required. Certainly, any subset of the left-hand side of \cref{eq:equality}
  (e.g., $\mathfrak{M}(p)$, where $\mathfrak{M}$ is a model of $(\phi, \sigma)$)
  is a subset of the right-hand side and vice versa. Since $\phi$ and $\psi$
  semantically only differ in what domains they refer to, every model of $\phi$
  satisfies $\psi$ and vice versa, completing the proof.
\end{proof}

@@

\begin{theorem}
  Any collection of functions returned by \textsc{Crane} can be evaluated in
  time polynomial in domain sizes.
\end{theorem}
\begin{proof}
  % 1) bound the number of possible arguments that any function can have
  % 2) prove that the computation time of any dynamic programming cell is always polytime (what about functions depending on other functions?))
  \begin{claim}
    Infinite recursion is impossible because that would necessitate a sequence
    of compilation rule applications that start with $\phi$ and end up with
    exactly $\phi$ as one of the outcomes. This is impossible because... [argue
    about formula length and the necessity to preserve domains and constants].
  \end{claim}
  \begin{claim}
    Argument values cannot increase from one function call to another.
  \end{claim}
  \begin{claim}
    Excluding functions, time to evaluate a function definition is polynomial
    (e.g., binomials, linear sums, powers, multiplication, addition). The number
    of operations within a definition is a constant that does not depend on
    domain sizes.
  \end{claim}
  \begin{claim}
    Describe a dynamic programming algorithm that:
    \begin{itemize}
      \item Gradually increases the sum of all domain sizes
      \item Computes and stores the values of all function calls whose arguments
            have that sum. Argue that there will always be an order in which
            that's possible.
    \end{itemize}
  \end{claim}
  NOTE: For proving that each cell is evaluated in polynomial time, I can just
  cite the ForcLift completeness (and domain recursion) paper.
\end{proof}

\section{Empirical Results}\label{sec:results}

TODO: Incidentally, the linear order axiom is also liftable. Cite the same
paper~\cite{DBLP:journals/corr/abs-2211-01164} in the introduction as well.

@@
